# Diagnostics experiments

## Diversity 

### Collective Fitness
Our initial investigations under the diversity diagnostic show that auxiliary age-driven diversity can be counterproductive in lexicase selection.

Treatments involving random injections with an age test case create oscillations in collective fitness. With every injection (which occurs every 
50,000 generations), collective fitness increases before decreasing below the baseline established by standard lexicase. This indicates that we are 
gaining temporary diversity but also losing older lineages that specialize in certain niches or subsets of traits.

Previous work by Dolson et al. shows that, under lexicase selection, lineages must occupy at least a minimal proportion of niches to avoid extinction, 
with this threshold decreasing as the population grows [CITE]. Our current configuration, which strongly selects for younger lineages by ensuring 
the age case appears early in a selection event, may be reducing the number of niches that older lineages could specialize in. Given the importance of
occupying niches for long-term survival, this further increases the likelihood of extinction for older lineages. 

Interestingly, random injections without age selection may degrade population-wide performance at higher population sizes. This effect becomes more 
prominent with larger injection sizes and occurs regardless of whether we are introducing a bottleneck event, that is, replacing a portion of the 
current population with injected individuals. At a population size of 500, injections without the age test case add no benefit but seem to remain 
stable and perform no worse than control in terms of collective fitness. Meanwhile, at a population size of 1,000, collective fitness increases 
slightly before degrading with each subsequent injection. The lack of effect at a population size of 500 might be due to the higher threshold that 
lineages must fulfill by occupying niches. Without selecting for age, new injects are unlikely to perform well enough on specific subsets of the 
problem, leading to their swift extinction. Meanwhile, the higher population size of 1,000, which results in a lower threshold, makes it more likely 
for new lineages to survive long enough to affect the overall fitness of the population.

While a lower threshold may prolong the survival of new injects, it is still unclear why degradation would occur without an age test case. 
Without selecting for age, it is unlikely for new lineages to be competitive against established specialists, particularly within their respective 
niches. Despite this, they seem to be displacing some older lineages, driving the latter to extinction. Suppose we consider the scenario that new 
injects are outcompeting older ones in their niches. This could cause the range of niches occupied by older lineages to fall below the 
aforementioned threshold. However, these new lineages somehow fail to establish long-term stability, perhaps due to an insufficient niche range 
or other factors, ultimately resulting in the degradation of collective fitness and loss of occupied niches or diversity.

### Lowest Active Position
There are no selective pressures for the lowest active position in the diversity diagnostic.

Mirroring the results for collective fitness, injections with an age case lead to oscillations in the lowest active position. With each injection, 
the lowest active position decreases before generally rising above the control runs after approximately 25,000 generations. 
Interestingly, the lowest active position rises much higher (beyond 20) for a population size of 500, but rises to about 6 for a population size 
of 1,000, regardless of cardinality.

Following the observations from collective fitness, we see that while the lowest active positions decrease due to increased diversity, we are 
simultaneously losing older lineages in the process. However, the loss of older lineages alone doesn't explain the frequent rise above the "baselines", as this 
should not visibly affect the lowest active positions if they are primarily driven by the new lineages. This rise might indicate that the new 
lineages are not performing well enough on the problem to survive long-term. Combined with the loss of older lineages, this could result in the 
lowest active positions being much higher than the baseline.

Injections without an age case at a lower population size of 500 seem to have no effect on the lowest active position. 
However, at a population size of 1,000, injections do impact the lowest active position of the population.

### Maximum Fitness
Random injections with an age case briefly lower maximum fitness, though the effect is virtually unnoticeable or insignificant at a population 
size of 1,000. No effect is observed for runs with random injections that do not include an age case.

## Exploit
We expected age-based diversity to be unhelpful in this diagnostic, but its performance is on par with the control in terms of both collective 
and maximum fitness.

The trends in collective fitness and the lowest active position seem to mirror each other. Initially, collective fitness is higher due to a more 
varied population. However, one or a few genotypes eventually dominates around the 25,000-generation mark, causing collective fitness to plateau. Random 
injections with an age test case briefly reintroduce some diversity, but the dominant lineage eventually takes over again.

Similarly, many individuals coexist early on, each focusing on different subsets of traits. By chance, some individuals have their maximal trait 
near the beginning of their genotype, leading to a lower lowest active position. Over time, one genotype dominates and 
outcompetes the rest. Since the position of the maximal trait is random, it averages around 50. Selecting for age temporarily decreases this 
value, but it rises again as new injects are overtaken by the dominant lineage.

As in the diversity diagnostic, there’s no inherent reason for the maximal value to be at any specific position in the genotype.

## Explore
Random injections with age selection seem to perform slightly better in terms of lowering the active position, but more work is needed to 
determine whether this improvement is significant. Age-based diversity also appears to slightly reduce or hinder maximum fitness, particularly in a 
smaller population of 500.

## Setup

### Dependencies
```{r}
library(ggplot2)
library(dplyr)
library(rlang)
```

### Set working directory
```{r}
experiment_slug <- "2024-07-06-diagnostics"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)

if (exists("bookdown_wd_prefix")) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}

data_directory <- paste0(working_directory, "data/")
average_directory <- paste0(data_directory, "average/")
```

### Plot averages
Due to the large number of plots, they cannot be displayed directly. However, they can be viewed in the average directory.
 
```{r}
generate_plot <- function(file_name, metric_name) {
  print(paste("Processing file:", file_name, "with metric:", metric_name))
  og_data <- read.csv(file_name, header=TRUE)
  
  # Making sure file_name is basename, remove extension: "diversity_200_500_100"
  file_base <- tools::file_path_sans_ext(basename(file_name))
  # Extract parameters from file_name: ["diversity", "200", "500", "100"]
  file_parts <- strsplit(file_base, "_")[[1]]
  type <- file_parts[1]
  num_vals <- file_parts[2]
  pop_size <- file_parts[3]
  inject_count <- file_parts[4]

  column_name <- ""
  if (metric_name == "Lowest Active Position") {
    column_name <- "Lowest.Active.Pos"
  }
  else if (metric_name == "Maximum Fitness") {
    column_name <- "Maximum.Fitness"
  }
  else if (metric_name == "Collective Fitness") {
    column_name <- "Collective.Fitness"
  }
  else {
    column_name <- metric_name
  }
  
  # Make sure the necessary columns exist
  required_columns <- c("X", "Type", column_name)
  if (!all(required_columns %in% colnames(og_data))) {
    stop(paste("One or more required columns are missing in the file:", file_name))
  }
  
  # Identify extra headers
  extra_header_rows <- which(grepl("Average Fitness", og_data$Average.Fitness))
  # Remove all extra headers
  clean_data <- og_data[-extra_header_rows, ]

  # Convert target column to numeric
  clean_data[[column_name]] <- as.numeric(clean_data[[column_name]])
  
  
  plot <- ggplot(clean_data, aes(x = X, y = !!sym(column_name), group = Type, color = Type)) +
    geom_line() +
    labs(
      title = paste("Averaged", metric_name, "Over Time"),
      subtitle = paste0("Experiment '", type, "', Cardinality ", num_vals, ", Pop. Size ", pop_size, ", Inject Size ", inject_count),
      x = "Generations (x1000)",
      y = metric_name
    )+
    theme_minimal()
  
  # Save the plot
  metric_name_file <- gsub(" ", "_", metric_name)
  
  output_filename <- paste0(tools::file_path_sans_ext(file_name), "_", metric_name_file, "_plot.png")
  ggsave(output_filename, plot=plot, width=10, height=6, bg="white")
}

metrics <- c("Maximum Fitness", "Collective Fitness", "Lowest Active Position")
all_files <- list.files(path=average_directory, pattern="*.csv")

# Exclude files starting with "C"
all_files <- all_files[!grepl("^C", all_files)]
print(all_files)

for (file in all_files) {
  for (metric in metrics) {
    file_path <- paste0(average_directory, file)
    generate_plot(file_path, metric)
  }
}
```