# Program synthesis experiments

```{r}
experiment_slug <- "2024-05-02-age-lex"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)

if (exists("bookdown_wd_prefix")) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}
```

## Dependencies

```{r}
library(tidyverse)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
library(kableExtra)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

### Load summary data

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

summary_data <- summary_data %>%
  mutate(
    PROBLEM = as.factor(PROBLEM),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    AGE_LEX_AGE_ORDER_LIMIT = as.factor(AGE_LEX_AGE_ORDER_LIMIT),
    RECOMB_PER_FUNC_SEQ_RECOMB_RATE = as.factor(RECOMB_PER_FUNC_SEQ_RECOMB_RATE),
    ORG_INJECTION_COUNT = as.factor(ORG_INJECTION_COUNT),
    ORG_INJECTION_MODE = factor(
      ORG_INJECTION_MODE,
      levels = c(
        "none",
        "random",
        "recombine-random",
        "recombine-complement"
      )
    ),
    ORG_INJECTION_INTERVAL = as.factor(ORG_INJECTION_INTERVAL),
    .keep = "all"
  )

solution_counts <- summary_data %>%
  group_by(
    PROBLEM,
    SELECTION,
    AGE_LEX_AGE_ORDER_LIMIT,
    ORG_INJECTION_COUNT,
    ORG_INJECTION_MODE,
    ORG_INJECTION_INTERVAL
  ) %>%
  summarize(
    solution_count = sum(found_solution == "1"),
    replicates = n(),
    no_solution_count = n() - sum(found_solution == "1"),
    elite_from_injected = sum(elite_elite_age != update),
    sol_from_injected = sum(elite_elite_age != update & found_solution == "1")
  )
# print(solution_counts, n=208)
solution_table <- kable(solution_counts) %>%
  kable_styling(latex_options = "striped", font_size = 25)
save_kable(solution_table, paste0(plot_directory, "solution_counts_table.pdf"))
solution_table
```

Solution by 8,000 updates?

```{r}
solution_counts_8k <- summary_data %>%
  group_by(
    PROBLEM,
    SELECTION,
    AGE_LEX_AGE_ORDER_LIMIT,
    ORG_INJECTION_COUNT,
    ORG_INJECTION_MODE,
    ORG_INJECTION_INTERVAL
  ) %>%
  summarize(
    solution_count = sum(found_solution == "1" & update <= 8000),
    replicates = n(),
    no_solution_count = n() - sum(found_solution == "1" & update <= 8000),
    # elite_from_injected = sum(elite_elite_age != update),
    sol_from_injected = sum((elite_elite_age != update) & (found_solution == "1" & update <= 8000))
  )
# print(solution_counts, n=208)
solution_table_8k <- kable(solution_counts_8k) %>%
  kable_styling(latex_options = "striped", font_size = 25)
save_kable(solution_table_8k, paste0(plot_directory, "solution_counts_table_8k.pdf"))
```


### Success rates across injection intervals

5k

```{r}
no_inject_sol <- solution_counts_8k %>%
  filter(ORG_INJECTION_MODE == "none") %>%
  ungroup() %>%
  select(!ORG_INJECTION_MODE)


plot <- solution_counts_8k %>%
  filter(ORG_INJECTION_MODE != "none") %>%
  ggplot(
    aes(
      x = ORG_INJECTION_INTERVAL,
      y = solution_count,
      fill = ORG_INJECTION_INTERVAL
    )
  ) +
  geom_col() +
  geom_text(
    aes(y = -0.5, label = solution_count)
  ) +
  geom_hline(
    data = no_inject_sol,
    aes(yintercept = solution_count),
    linetype = "dashed"
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  scale_x_discrete(
    name = "Injection Interval"
  ) +
  scale_y_continuous(
    "Solution count"
  ) +
  facet_grid(
    ORG_INJECTION_MODE ~ PROBLEM
  ) +
  theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )
ggsave(
  filename = paste0(plot_directory, "solutions-bar-8k.pdf"),
  plot = plot,
  width = 15,
  height = 10
)
```

### Injection modes at interval 100

```{r}
plot <- solution_counts_8k %>%
  # filter(ORG_INJECTION_INTERVAL == "200" | ORG_INJECTION_MODE == "none") %>%
  # filter(ORG_INJECTION_MODE == "none") %>%
  ggplot(
    aes(
      x = ORG_INJECTION_MODE,
      y = solution_count,
      fill = ORG_INJECTION_MODE
    )
  ) +
  geom_col() +
  geom_text(
    aes(y = -0.5, label = solution_count)
  ) +
  # geom_hline(
  #   data = no_inject_sol,
  #   aes(yintercept = solution_count),
  #   linetype = "dashed"
  # ) +
  scale_fill_light() +
  scale_color_light() +
  scale_x_discrete(
    name = "Injection Mode",
    breaks = c("none", "random", "recombine-random", "recombine-complement"),
    labels = c("None", "Rand", "Recomb-Rand", "Recomb-Comp")
  ) +
  scale_y_continuous(
    "Solution count"
  ) +
  facet_grid(
    ORG_INJECTION_INTERVAL ~ PROBLEM
    # nrow = 3
  ) +
  theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )
ggsave(
  filename = paste0(plot_directory, "solutions-interalall-8k.pdf"),
  plot = plot,
  width = 15,
  height = 5
)
```

#### Statistics

```{r}
sol_stats_data <- solution_counts_8k %>%
  filter(ORG_INJECTION_INTERVAL == "100" | ORG_INJECTION_MODE == "none") %>%
  ungroup() %>%
  select(
    PROBLEM, ORG_INJECTION_MODE, solution_count, no_solution_count
  )
```

```{r}
fisher_results <- data.frame(
  comparison = character(),
  group1 = character(),
  group2 = character(),
  n = integer(),
  p = double(),
  p.adj = double(),
  p.adj.signif = character()
)

problems <- levels(sol_stats_data$PROBLEM)
for (problem in problems) {

  ft_results <- sol_stats_data %>%
    filter(PROBLEM == problem) %>%
    select(!PROBLEM) %>%
    column_to_rownames(var = "ORG_INJECTION_MODE") %>%
    pairwise_fisher_test(
      p.adjust.method = "holm"
    ) %>%
    add_significance("p.adj")

  ft_results <- ft_results %>%
    mutate(
      comparison = rep(problem, nrow(ft_results)),
      .keep = "all"
    ) %>%
    relocate(comparison)

  fisher_results <- rbind(
    fisher_results,
    ft_results
  )
}
fisher_results <- as.tibble(fisher_results)
fisher_results <- fisher_results %>%
  mutate(
    comparison = as.factor(comparison),
    group1 = as.factor(group1),
    group2 = as.factor(group2),
  ) %>%
  group_by(
    comparison
  )

fisher_table <- kbl(fisher_results) %>% kable_styling()
save_kable(fisher_table, paste0(plot_directory, "stats_table_int100_8k.pdf"))
fisher_table
```

### Elite from original population or injected

Solutions from injected

```{r}
# summary_data %>%
#   pivot_longer(
#     cols =
#   )

plot <- solution_counts_8k %>%
  filter(ORG_INJECTION_INTERVAL == "100" | ORG_INJECTION_MODE == "none") %>%
  ggplot(
    aes(
      x = ORG_INJECTION_MODE,
      y = sol_from_injected,
      fill = ORG_INJECTION_MODE
    )
  ) +
  geom_col(aes(y = solution_count), fill = "grey", alpha = 0.8) +
  geom_col() +
  geom_text(
    data = filter(solution_counts_8k, (ORG_INJECTION_INTERVAL == "100" | ORG_INJECTION_MODE == "none") & solution_count > 0),
    aes(
      # y = sol_from_injected,
      y = -0.5,
      label = round(sol_from_injected / solution_count, 2)
    )
    # vjust = 1.2
  ) +
  scale_fill_light() +
  scale_color_light() +
  scale_x_discrete(
    name = "Injection Mode",
    breaks = c("none", "random", "recombine-random", "recombine-complement"),
    labels = c("None", "Rand", "Recomb-Rand", "Recomb-Comp")
  ) +
  scale_y_continuous(
    "Solution count"
  ) +
  facet_wrap(
    ~ PROBLEM,
    nrow = 1
  ) +
  theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )
ggsave(
  filename = paste0(plot_directory, "sol-from-inject-interal100-8k.pdf"),
  plot = plot,
  width = 15,
  height = 5
)
```

### Elite age

Note this is not a particularly meaningful plot because solutions (that stop early) are mixed in.
Not sure if elite age is driven by solution stop or because solution's lineage age.

```{r}
# elite_age
plot <- summary_data %>%
  filter(
    SELECTION == "age-lexicase" & ORG_INJECTION_INTERVAL == "100"
  ) %>%
  ggplot(
    aes(
      x = ORG_INJECTION_MODE,
      y = elite_elite_age / update,
      fill = ORG_INJECTION_MODE
    )
  ) +
  # geom_flat_violin(
  #   position = position_nudge(x = .2, y = 0),
  #   alpha = .8,
  #   adjust = 1.5
  # ) +
  geom_point(
    mapping = aes(color = ORG_INJECTION_MODE),
    position = position_jitter(width = .2),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .2,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    name = "Elite age (as % of total generations)"
  ) +
  scale_x_discrete(
    name = "Injection Mode",
    breaks = c("none", "random", "recombine-random", "recombine-complement"),
    labels = c("None", "Rand", "Recomb-Rand", "Recomb-Comp")
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    ~PROBLEM,
    nrow = 1
    # ncol = 1
  ) +
  # coord_flip() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )
ggsave(
  filename = paste0(plot_directory, "elite-age.pdf"),
  plot = plot,
  width = 15,
  height = 5
)
```

Statistics

We run a Kruskal-Wallis test to check for any differences in distributions.

```{r}
focal_data <- summary_data %>%
  filter(
    SELECTION == "age-lexicase" & ORG_INJECTION_INTERVAL == "100"
  ) %>%
  mutate(
    age_proportion = elite_elite_age / update
  )


kw_test <- focal_data %>%
  group_by(SELECTION, PROBLEM) %>%
  kruskal_test(age_proportion ~ ORG_INJECTION_MODE) %>%
  mutate(sig = (p < 0.05))
  # unite(
  #   "comparison_group",
  #   SELECTION,
  #   evals_per_gen,
  #   sep = "_",
  #   remove = FALSE
  # )
kable(kw_test)
```

```{r}
# # Grab group names of significant comparisons
sig_kw_groups <- filter(kw_test, p < 0.05)$PROBLEM

wrs_test <- focal_data %>%
  filter(
    PROBLEM %in% sig_kw_groups
  ) %>%
  group_by(SELECTION, PROBLEM) %>%
  pairwise_wilcox_test(age_proportion ~ ORG_INJECTION_MODE) %>%
  adjust_pvalue(method = "holm") %>%
  add_significance("p.adj")
wrs_table <- kbl(wrs_test) %>% kable_styling()
save_kable(wrs_table, paste0(plot_directory, "age_stats_table.pdf"))

# kable(wrs_test)
```