# Diagnostics configuration

In addition to the GP and Sudoku experiments, we tested age-based lexicase against the exploration, exploitation, and diversity diagnostic 
fitness landscapes included in the MABE 2.0 software framework. With the exception of the diversity diagnostic, both exploration and exploitation 
are detailed in Hernandez et al. (2018) [CITE]. The diversity diagnostic is similar to the "Contradictory Objectives" test in the paper but 
translates the inactive genes to the phenotype by subtracting them from the active (maximal) trait and then halving that value. 
This creates pressure to maximize one trait while minimizing the rest. In accordance with Hernandez et al., we used a genotype representation 
consisting of a sequence of floating-point values.

We measured:
- The **maximum fitness** attained in the population
- The sum of the highest values for each trait across the population (**collective fitness**)
- The current lowest position of maximal trait in the population (**lowest "active" position**)

Below, we document the configurations used for these experiments.

## MABE configuration files

```{r}
experiment_slug <- "2024-07-06-diagnostics"

config_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/hpc/",
  "/config/"
)

if (exists("bookdown_wd_prefix")) {
  config_directory <- paste0(
    bookdown_wd_prefix,
    config_directory
  )
}
```

### Standard Lexicase
```{r comment=''}
file <- paste0(config_directory, "/AgeControl-No-Inject.mabe")
file_content <- readLines(file)
cat(file_content, sep='\n')
```

### Standard Lexicase with Random Injection
```{r comment=''}
file <- paste0(config_directory, "/AgeControl.mabe")
file_content <- readLines(file)
cat(file_content, sep='\n')
```

### Standard Lexicase with Random Injection and No Replacement
```{r comment=''}
file <- paste0(config_directory, "/AgeControl-No-Replace.mabe")
file_content <- readLines(file)
cat(file_content, sep='\n')
```

### Age-Based Lexicase
```{r comment=''}
file <- paste0(config_directory, "/AgeDiagnostics.mabe")
file_content <- readLines(file)
cat(file_content, sep='\n')
```

### Age-Based Lexicase with No Replacement
```{r comment=''}
file <- paste0(config_directory, "/AgeDiagnostics-No-Replace.mabe")
file_content <- readLines(file)
cat(file_content, sep='\n')
```


## Experiment conditions
We varied the cardinality (the number of values for a genotype, or "num_vals"), population size, and inject count, 
while keeping the number of generations and injection interval consistent throughout the experiments.

We conducted exploit experiments exclusively for AgeControl.mabe (standard lexicase with random injection) and AgeDiagnostics.mabe (age-based lexicase).

Each condition was tested with 20 replicates across all experiments.

```
fixed_parameters = {
    "num_gens":"1000000",
    "print_step":"1000",
    "inject_step":"50000",
    "elite_count":"0",
}

special_decorators = ["__DYNAMIC", "__COPY_OVER"]

# Register parameters that vary across treatments/runs
combos.register_var("filename")
combos.register_var("diagnostic__COPY_OVER")
combos.register_var("pop_inject_ratio__COPY_OVER")
combos.register_var("num_vals__COPY_OVER")

combos.add_val(
    "filename",
    [
        "AgeControl.mabe",
        "AgeDiagnostics.mabe"
        "AgeControl-No-Inject.mabe",
        "AgeControl-No-Replace.mabe",
        "AgeDiagnostics-No-Replace.mabe"
    ]
)

combos.add_val(
    "diagnostic__COPY_OVER",
    [
        "-s diagnostics.diagnostic=\\\"explore\\\"",
        "-s diagnostics.diagnostic=\\\"diversity\\\""
    ]
)

combos.add_val(
    "pop_inject_ratio__COPY_OVER",
    [
        "-s pop_size=500 -s inject_count=50",
        "-s pop_size=1000 -s inject_count=100",
        "-s pop_size=500 -s inject_count=100",
        "-s pop_size=1000 -s inject_count=200",
        "-s pop_size=500 -s inject_count=200",
        "-s pop_size=1000 -s inject_count=400"
    ]
)
combos.add_val(
    "num_vals__COPY_OVER",
    [
        "-s num_vals=200",
        "-s num_vals=500",        
    ]
)
```

```
fixed_parameters = {
    "num_gens":"100000",
    "print_step":"1000",
    "inject_step":"50000",
    "elite_count":"0",
}

special_decorators = ["__DYNAMIC", "__COPY_OVER"]

# Register parameters that vary across treatments/runs
combos.register_var("filename")
combos.register_var("diagnostic__COPY_OVER")
combos.register_var("pop_inject_ratio__COPY_OVER")
combos.register_var("num_vals__COPY_OVER")

combos.add_val(
    "filename",
    [
        "AgeControl.mabe",
        "AgeDiagnostics.mabe"
    ]
)

combos.add_val(
    "diagnostic__COPY_OVER",
    [
        "-s diagnostics.diagnostic=\\\"exploit\\\"",
    ]
)

combos.add_val(
    "pop_inject_ratio__COPY_OVER",
    [
        "-s pop_size=500 -s inject_count=50",
        "-s pop_size=1000 -s inject_count=100",
        "-s pop_size=500 -s inject_count=100",
        "-s pop_size=1000 -s inject_count=200",
        "-s pop_size=500 -s inject_count=200",
        "-s pop_size=1000 -s inject_count=400"
    ]
)
combos.add_val(
    "num_vals__COPY_OVER",
    [
        "-s num_vals=200",
        "-s num_vals=500",        
    ]
)
```
